Synthesising Complex Termination Arguments

### 2014-06-20 papers/termination/paper.tex
Some more work on section 6

We can specify the existence of a ranking function, and therefore the termination
of a program, with the second order formulae defined in Fig.~\ref{fig:termination-formulae}.
We can observe that each of these formulae (with the exception of Definition~\ref{def:nonterm-formula})
is an $\exists \forall$ formula where the the universally quantified variables only range over
ground terms.  This class of formulae is the same as the class of program synthesis formulae.
The program synthesis problem can be described as finding a satisfying assigment to the
synthesis formula:

\begin{definition}[Synthesis Formula]
 \label{def:synth-formula}
 $$\exists \vec{P}, \vec{x} \cdotp \forall \vec{y} \cdotp \sigma(\vec{x}, \vec{P}, \vec{y}) $$
\end{definition}

Since this formula involves quantification over functions $X \to Y$,
this is a second-order formula -- indeed, if $X = Y = \mathbb{N}$,
the formula describes a set at level $\Sigma_1^1$ of the analytical
hierarchy.  As such, determining the satisfiability of the synthesis
formula over a given logic is strictly harder than solving the
halting problem over the same logic.

\subsection{Reducing Termination to Synthesis}
For a transition system $T$, we can build a specification $\sigma$ to find a conditional ranking function for
$T$ with initial states I.  This specification is defined in Definition~\ref{def:termination-spec}.

Here we have synthesised a ranking function $R$ into $\mathbb{N}$ (which is well-founded),
as well as an inductive invariant $V$ that guarantees termination of $T$.

To synthesise an n-place lexicographic ranking function, we just ask for a ranking function
$R: X \to \mathbb{N}^n$.

We recall the definition of the Kolmogorov-complexity of a function $f$:

@@

We are primarily interested in studying low-Kolmogorov-complexity (LKC)
functions.

\begin{theorem}
 Linear functions are LKC.
\end{theorem}

\begin{proof}
 The program to compute a linear function $f: X \to Y$ is of size linear in
 $\dim(X) \times \dim(Y)$.
\end{proof}

\begin{conjecture}
 Most LKC functions are non-linear.
\end{conjecture}

>>>>

We can specify the existence of a ranking function, and therefore the termination
of a program, with the second order formulae defined in Fig.~\ref{fig:termination-formulae}.
We can observe that each of these formulae (with the exception of Definition~\ref{def:nonterm-formula})
is an $\exists \forall$ formula where the the universally quantified variables only range over
ground terms.
The program synthesis problem can be described as finding a satisfying assigment to the
synthesis formula:

\begin{definition}[Synthesis Formula]
 \label{def:synth-formula}
 $$\exists P, x \cdot \forall y \cdot \sigma(x, y, P)$$
\end{definition}

Here we can view $\sigma: X \times Y \times (X \times Y \to Z) \to \mathbb{B}$ as a specification function.
The idea is that $\sigma$ returns true iff the function $P$ computes appropriate outputs
when fed the inputs $x$ and $y$.  If we can find a function $P$ and a value $x$ that meet the specification
for all $y$, we have found a solution to the synthesis problem and write $P \models \sigma$.  Since the formulae
of Fig.~\ref{fig:termination-formulae} are syntactically in the synthesis fragment, by concocting an appropriate
specification function we reduce the problem of proving termination to the problem of
program synthesis.

\subsection{Synthesis Complexity}
In Section~\ref{sec:synthesis} we will describe a procedure for solving the synthesis formula.
This procedure has the property that it will find the shortest program meeting the
specification, and as a consequence the runtime of the program synthesiser is dominated
by the size of this shortest program.  To frame the discussion of the runtime of our
termination proving procedure, we first recall the definition of the Kolmogorov complexity
of a function $f$:

@@

We can extend this definition slightly to talk about the Kolmogorov complexity of a
synthesis problem in terms of its specification:

\begin{definition}[Kolmogorov complexity of a synthesis problem]
 The Kolmogorov complexity of a program specification $K(\sigma)$ is the length of the shortest
 program $P$ such that $P \models \sigma$.
\end{definition}

We will later show that the time complexity of our synthesis procedure is $NP^{NP}(K(\sigma))$,
which means that for even moderately large $K(\sigma)$ our procedure is intractably slow.  Conversely
if $K(\sigma)$ is small we will be able to find a termination proof, regardless of other parameters
such as the size of the program under analysis.  To put this another way, under the assumption
that $K(\sigma)$ is small for a program's termination problem, we will find a ranking function
quite rapidly.  We will now investigate the properties of functions with low Kolmogorov complexity,
and show that the assumption of a low-Kolmogorov ranking function is much weaker than the assumption
of a linear ranking function.

\begin{theorem}
 Linear functions have low Kolmogorov complexity.
\end{theorem}

\begin{proof}
 The function $f: X \to Y$ can be computed with a program consisting of
 $2 \cdot \dim(X) \cdot \dim(Y) - \dim(Y)$ instructions (1 multiplication and 1 addition for
 each cell in the matrix representing $f$).  Therefore,
 $K(f) \leq 2 \cdot \dim(X) \cdot \dim(Y) - \dim(Y)$.
\end{proof}

\begin{theorem}
 Most functions are non-linear.
\end{theorem}
